Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-14 talk/icooolps2011/paper.tex
write acknowledgements
ruthlessly cut space, mostly by making font smaller. not pretty, but I have no idea what to cut atm.

\usepackage[T1]{fontenc}
\usepackage[scaled=0.8]{beramono}

@@

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010} and a Smalltalk VM
\cite{carl_friedrich_bolz_back_2008}.

@@

\begin{figure}
\includegraphics[scale=0.5]{figures/trace-levels}
\caption{The levels involved in tracing}
\label{fig:trace-levels}
\end{figure}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
if x == 4:
    y = y + x
\end{lstlisting}

If the fragment is traced with $x_1$ being \texttt{4}, the following trace is
produced:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$y_2$ = $y_1$ + $x_1$
\end{lstlisting}

@@

Let's make this more concrete. If we trace a call to the following function:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_2$ = 9 + $y_1$
return($v_2$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 2)
$v_2$ = 5 + $y_1$
return($v_2$)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$x_1$ = $a_1$.x
$v_1$ = $x_1$ * 2
$v_2$ = $v_1$ + 1
$v_3$ = $v_2$ + $val_1$
$a_1$.y = $v_3$
\end{lstlisting}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    @elidable
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

Now the trace will look like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_1$ = compute($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

\begin{figure}
\begin{center}
{\footnotesize
\begin{tabular}{|l|r|r|r|}
\hline
	&CPython	&JIT baseline	&JIT full\\
\hline
django[ms]	&988.67 $\pm$ 0.49	&405.62 $\pm$ 4.80	&149.31 $\pm$ 1.37\\
	&6.62 $\times$	&2.72 $\times$	&1.00 $\times$\\
\hline
go[ms]	&947.43 $\pm$ 1.30	&525.53 $\pm$ 7.67	&174.32 $\pm$ 7.78\\
	&5.44 $\times$	&3.01 $\times$	&1.00 $\times$\\
\hline
pyflate[ms]	&3209.20 $\pm$ 3.65	&2884.26 $\pm$ 21.11	&1585.48 $\pm$ 5.22\\
	&2.02 $\times$	&1.82 $\times$	&1.00 $\times$\\
\hline
richards[ms]	&357.79 $\pm$ 1.32	&421.87 $\pm$ 0.48	&17.89 $\pm$ 1.15\\
	&20.00 $\times$	&23.58 $\times$	&1.00 $\times$\\
\hline
telco[ms]	&1209.67 $\pm$ 2.20	&738.18 $\pm$ 3.29	&153.48 $\pm$ 1.86\\
	&7.88 $\times$	&4.81 $\times$	&1.00 $\times$\\
\hline
\end{tabular}
}
\end{center}
\caption{Benchmark Results}
\label{fig:times}
\end{figure}

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
\cite{carl_friedrich_bolz_towards_????}. Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion was also used in DyC \cite{grant_dyc:_2000}, a
runtime partial evaluator for C. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

@@

XXX Peng Wu and David Edelsohn, Laura Creighton

>>>>

\usepackage[T1]{fontenc}
\usepackage[scaled=0.82]{beramono}

@@

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010}.

@@

\begin{figure}
\includegraphics[scale=0.45]{figures/trace-levels}
\caption{The levels involved in tracing}
\label{fig:trace-levels}
\end{figure}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
if x == 4:
    y = y + x
\end{lstlisting}
}

If the fragment is traced with $x_1$ being \texttt{4}, the following trace is
produced:
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$y_2$ = $y_1$ + $x_1$
\end{lstlisting}
}

@@

Let's make this more concrete. If we trace a call to the following function:
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_2$ = 9 + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 2)
$v_2$ = 5 + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}
}

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$x_1$ = $a_1$.x
$v_1$ = $x_1$ * 2
$v_2$ = $v_1$ + 1
$v_3$ = $v_2$ + $val_1$
$a_1$.y = $v_3$
\end{lstlisting}
}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    @elidable
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}
}

Now the trace will look like this:
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_1$ = compute($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

@@

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|}
\hline
	&CPython	&JIT baseline	&JIT full\\
\hline
django[ms]	&988.67 $\pm$ 0.49	&405.62 $\pm$ 4.80	&149.31 $\pm$ 1.37\\
	&6.62 $\times$	&2.72 $\times$	&1.00 $\times$\\
\hline
go[ms]	&947.43 $\pm$ 1.30	&525.53 $\pm$ 7.67	&174.32 $\pm$ 7.78\\
	&5.44 $\times$	&3.01 $\times$	&1.00 $\times$\\
\hline
pyflate[ms]	&3209.20 $\pm$ 3.65	&2884.26 $\pm$ 21.11	&1585.48 $\pm$ 5.22\\
	&2.02 $\times$	&1.82 $\times$	&1.00 $\times$\\
\hline
richards[ms]	&357.79 $\pm$ 1.32	&421.87 $\pm$ 0.48	&17.89 $\pm$ 1.15\\
	&20.00 $\times$	&23.58 $\times$	&1.00 $\times$\\
\hline
telco[ms]	&1209.67 $\pm$ 2.20	&738.18 $\pm$ 3.29	&153.48 $\pm$ 1.86\\
	&7.88 $\times$	&4.81 $\times$	&1.00 $\times$\\
\hline
\end{tabular}
}
\end{center}
\caption{Benchmark Results}
\label{fig:times}
\end{figure}

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT as well as a Prolog partial evaluator
\cite{bolz_towards_2009}. Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion was also used in DyC \cite{grant_dyc:_2000}, a
runtime partial evaluator for C. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

@@

The authors would like to thank Peng Wu, David Edelsohn and Laura Creighton for
encouragement, fruitful discussions and feedback during the writing of this paper.
